Spatial Generalized Linear Models in \proglang{R} Using \pkg{spmodel}

### 2025-11-14 inst/manuscript/manuscript.Rmd
some updates

where $[\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]$ is the density for the appropriate response distribution of $\mathbf{y}$ (e.g., binomial, Poisson) given the latent $\mathbf{w}$ and dispersion parameter ($\varphi$), and $[\mathbf{w} | \mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\theta}]$ is the multivariate Gaussian density for $\mathbf{w}$ given the explanatory variables ($\mathbf{X}$), fixed effects ($\boldsymbol{\beta}$), and spatial covariance parameters ($\boldsymbol{\theta}$). The elements of $[\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]$ are conditionally independent (given $\mathbf{w}$), but the elements of $[\mathbf{w} | \mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\theta}]$ share spatial covariance. Following @harville1977maximum, we can integrate $\boldsymbol{\beta}$ out of Equation$~$\ref{eq-marginal}, which yields

@@

A second-order Taylor series expansion of $\ell_\mathbf{w}$ around a point $\hat{\mathbf{w}}$ yields

@@

  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \int_{\mathbf{w}} \exp(\ell_{\hat{\mathbf{w}}} + \mathbf{g}^\top(\mathbf{w} - \hat{\mathbf{w}}) + \frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^\top \mathbf{G} (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w},

@@

where $\mathbf{g}$ and $\mathbf{G}$ are the gradient and Hessian, respectively, of $\ell_\mathbf{w}$ with respect to $\mathbf{w}$. If $\hat{\mathbf{w}}$ is a value for which $\mathbf{g} = \mathbf{0}$,

@@

  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \exp(\ell_{\hat{\mathbf{w}}}) \int_{\mathbf{w}} \exp(-\frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^\top (-\mathbf{G}) (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w}.

@@

The integral in Equation$~$\ref{eq-marginal5} can be solved by leveraging properties of the normalizing constant of a multivariate Gaussian distribution. Thus, rewriting $\exp(\ell_{\hat{\mathbf{w}}})$ yields

@@

  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\hat{\mathbf{w}}), \varphi] [\hat{\mathbf{w}} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\hat{\mathbf{w}}}|^{-1/2}.

@@

Maximizing the natural logarithm of Equation$~$\ref{eq-marginal6} requires a doubly iterative process over $\boldsymbol{\theta}$ and $\varphi$ as well as $\mathbf{w}$, eventually yielding the the marginal restricted maximum likelihood estimators $\hat{\varphi}$ and $\hat{\boldsymbol{\theta}}$ and their corresponding values of $\hat{\mathbf{w}}$. Maximizing this log likelihood requires repeatedly evaluating $\boldsymbol{\Sigma}^{-1}$, $\mathbf{g}$, and $\mathbf{G}$; see @ver2024marginal for more details and comparisons to other approaches as well as forms of $\mathbf{g}$ and $\mathbf{G}$ for various response distributions.

@@

Though the fixed effects are integrated out of the likelihood, we can still estimate them using generalized least squares (GLS) principles, a common practice for linear models estimated using restricted maximum likelihood methods. Had we observed $\mathbf{w}$, a GLS estimator for $\boldsymbol{\beta}$ is given by
\begin{align}\label{eq-gls1}
  \hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{w} = \mathbf{B}\mathbf{w},
\end{align}
where $\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}$. However, we only observe $\hat{\mathbf{w}}$, so it is reasonable to define $\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}$. Thus, to derive properties of $\hat{\boldsymbol{\beta}}$ like expectation and variance, we must derive these properties for $\hat{\mathbf{w}}$. To do so, we must condition on $\mathbf{w}$ as if it were observed and invoke properties of the laws of total expectation and variance. Because $\hat{\mathbf{w}}$ was obtained by maximizing the likelihood, we may assume that given $\mathbf{w}$, $\hat{\mathbf{w}}$ has mean $\mathbf{w}$ and variance approximately equal to $-\mathbf{H}^{-1}$ (the inverse Hessian). It follows that $\text{E}(\hat{\mathbf{w}})$ is given by
\begin{align}
  \text{E}(\hat{\mathbf{w}}) = \text{E}(\text{E}(\hat{\mathbf{w}} | \mathbf{w}))
   = \text{E}(\mathbf{w}) 
   = \mathbf{X}\boldsymbol{\beta}
\end{align}
and $\text{Var}(\hat{\mathbf{w}})$ is given by
\begin{align}
  \text{Var}(\hat{\mathbf{w}}) & = \text{E}(\text{Var}(\hat{\mathbf{w}} | \mathbf{w})) + \text{Var}(\text{E}(\hat{\mathbf{w}} | \mathbf{w})) \\
  & = \text{E}(-\mathbf{H}^{-1}) + \text{Var}(\mathbf{w})\\
  & = -\mathbf{H}^{-1} + \boldsymbol{\Sigma}
\end{align}
Putting this all together, it follows that $\hat{\boldsymbol{\beta}}$ is unbiased for $\boldsymbol{\beta}$:
\begin{align}
  \text{E}(\hat{\boldsymbol{\beta}})  = \text{E}(\mathbf{B}\hat{\mathbf{w}}) 
  = \mathbf{B}\text{E}(\hat{\mathbf{w}}) =  (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})\boldsymbol{\beta} = \boldsymbol{\beta}.
\end{align}
Moreover, it follows that
\begin{align}
  \text{Var}(\hat{\boldsymbol{\beta}}) & = \text{Var}(\mathbf{B}\hat{\mathbf{w}}) \\
  & = \mathbf{B} \text{Var}(\hat{\mathbf{w}}) \mathbf{B}^\top\\
  & = \mathbf{B} (-\mathbf{H}^{-1} + \boldsymbol{\Sigma}) \mathbf{B}^\top \\
  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + \mathbf{B}\boldsymbol{\Sigma}\mathbf{B}^\top \\
  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}.
\end{align}
In practice, $\text{Var}(\hat{\boldsymbol{\beta}})$ is estimated by evaluating $\boldsymbol{\Sigma}$ at $\hat{\boldsymbol{\theta}}$, the estimated covariance parameter vector.

@@

Inspecting model diagnostics is an important step of the modeling process that can yield valuable insights into model behavior and unusual observations. @montgomery2021introduction contextualize three components of unusual observations: outliers, leverage, and influence. An observation is an outlier if it has an extreme response value relative to expectation. The response GLM residuals simply compare the observation to its fitted latent mean:
\begin{align}
  \mathbf{r}_{r} = \mathbf{y} - f^{-1}(\hat{\mathbf{w}})
\end{align}
Because observations often have a unique support in a GLM (e.g., only two possible response values for binary data) and the variance of an observation generally depends on its mean, response residuals lack some utility. Deviance residuals are a function of response residuals that are appropriately scaled to behave more like response residuals in a standard linear model. Deviance residuals are given by
\begin{align}
  \mathbf{r}_{d} = sign(\mathbf{r}_{r})\sqrt{\mathbf{d}},
\end{align}
where $\mathbf{d}$ is a vector of individual deviances. The sum of the squared deviance residuals equals the sum of $\mathbf{d}$. The sum of $\mathbf{d}$ is the deviance of the model fit, which quantifies twice the difference in log likelihoods between the a saturated model that fits every observation perfectly (i.e., $\mathbf{y} = f^{-1}(\hat{\mathbf{w}}_i)$ for all $i$) and the fitted model [@myers2012generalized]. Deviance is often used as a fit statistic; lower values of deviance imply a better model fit (compared to the observed data). Pearson and standardized residuals are other types of GLM residuals that involve a scaling of the response residuals; the Pearson residuals scale $\mathbf{r}_{r}$ by the square root of $\mathbf{V}$, while the standardized residuals scale the deviance residuals by $\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}$, where $\mathbf{L}_{ii}$ is the $i$th diagonal element of the leverage matrix, which we discuss next.

>>>>

where $[\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]$ is the density for the appropriate response distribution of $\mathbf{y}$ (e.g., binomial, Poisson) given the latent $\mathbf{w}$ and dispersion parameter ($\varphi$), and $[\mathbf{w} | \mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\theta}]$ is the multivariate Gaussian density for $\mathbf{w}$ given the explanatory variables ($\mathbf{X}$), fixed effects ($\boldsymbol{\beta}$), and spatial covariance parameters ($\boldsymbol{\theta}$). The elements of $[\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]$ are conditionally independent (given $\mathbf{w}$), but the elements of $[\mathbf{w} | \mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\theta}]$ share spatial covariance. Following @harville1974bayesian, we can integrate $\boldsymbol{\beta}$ out of Equation$~$\ref{eq-marginal}, which yields

@@

A second-order Taylor series expansion of $\ell_\mathbf{w}$ around a point $\mathbf{w}^{*}$ yields

@@

  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \int_{\mathbf{w}} \exp(\ell_{\mathbf{w}^{*}} + \mathbf{g}^\top(\mathbf{w} - \mathbf{w}^{*}) + \frac{1}{2}(\mathbf{w} - \mathbf{w}^{*})^\top \mathbf{G} (\mathbf{w} - \mathbf{w}^{*}))d\mathbf{w},

@@

where $\mathbf{g}$ and $\mathbf{G}$ are the gradient and Hessian, respectively, of $\ell_\mathbf{w}$ with respect to $\mathbf{w}$. If $\mathbf{w}^{*}$ is a value for which $\mathbf{g} = \mathbf{0}$,

@@

  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \exp(\ell_{\mathbf{w}^{*}}) \int_{\mathbf{w}} \exp(-\frac{1}{2}(\mathbf{w} - \mathbf{w}^{*})^\top (-\mathbf{G}) (\mathbf{w} - \mathbf{w}^{*}))d\mathbf{w}.

@@

The integral in Equation$~$\ref{eq-marginal5} can be solved by leveraging properties of the normalizing constant of a multivariate Gaussian distribution. Thus, rewriting $\exp(\ell_{\mathbf{w}^{*}})$ yields

@@

  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\mathbf{w}^{*}), \varphi] [\mathbf{w}^{*} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\mathbf{w}^{*}}|^{-1/2}.

@@

Maximizing the natural logarithm of Equation$~$\ref{eq-marginal6} requires a doubly iterative process over 1) $\boldsymbol{\theta}$ and $\varphi$ and 2) the latent $\mathbf{w}$ (to find each set of $\mathbf{w}^{*}$). This doubly iterative process eventually yields the marginal restricted maximum likelihood estimators $\hat{\varphi}$ and $\hat{\boldsymbol{\theta}}$ and their respective values of $\mathbf{w}^{*}$, which we call $\hat{\mathbf{w}}$. Maximizing the marginal restricted likelihood requires repeatedly evaluating $\boldsymbol{\Sigma}^{-1}$, $\mathbf{g}$, and $\mathbf{G}$; see @ver2024marginal for more details and comparisons to other approaches as well as forms of $\mathbf{g}$ and $\mathbf{G}$ for various response distributions.

@@

Though the fixed effects are integrated out of the likelihood, we can still estimate them using generalized least squares (GLS) principles, a common practice for linear models estimated using restricted maximum likelihood methods. Had we observed $\mathbf{w}$, a GLS estimator for $\boldsymbol{\beta}$ is given by
\begin{align}\label{eq-gls1}
  \hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{w} = \mathbf{B}\mathbf{w},
\end{align}
where $\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}$. However, we do not observe $\mathbf{w}$ and instead estimate $\mathbf{w}$ via $\hat{\mathbf{w}}$. Thus, it is reasonable to define $\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}$. To derive properties of $\hat{\boldsymbol{\beta}}$ like expectation and variance, we must derive these properties for $\hat{\mathbf{w}}$. To do so, we must condition on $\mathbf{w}$ as if it were observed and invoke properties of the laws of total expectation and variance. Because $\hat{\mathbf{w}}$ was obtained by maximizing the likelihood, we may assume that given $\mathbf{w}$, $\hat{\mathbf{w}}$ has mean $\mathbf{w}$ and variance approximately equal to $-\mathbf{H}^{-1}$ (the inverse Hessian). It follows that $\text{E}(\hat{\mathbf{w}})$ is given by
\begin{align}
  \text{E}(\hat{\mathbf{w}}) = \text{E}(\text{E}(\hat{\mathbf{w}} | \mathbf{w}))
   = \text{E}(\mathbf{w}) 
   = \mathbf{X}\boldsymbol{\beta}
\end{align}
and $\text{Var}(\hat{\mathbf{w}})$ is given by
\begin{align}
  \text{Var}(\hat{\mathbf{w}}) & = \text{E}(\text{Var}(\hat{\mathbf{w}} | \mathbf{w})) + \text{Var}(\text{E}(\hat{\mathbf{w}} | \mathbf{w})) \\
  & = \text{E}(-\mathbf{H}^{-1}) + \text{Var}(\mathbf{w})\\
  & = -\mathbf{H}^{-1} + \boldsymbol{\Sigma}
\end{align}
Putting this all together, it follows that $\hat{\boldsymbol{\beta}}$ is unbiased for $\boldsymbol{\beta}$:
\begin{align}
  \text{E}(\hat{\boldsymbol{\beta}})  = \text{E}(\mathbf{B}\hat{\mathbf{w}}) 
  = \mathbf{B}\text{E}(\hat{\mathbf{w}}) =  (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})\boldsymbol{\beta} = \boldsymbol{\beta}.
\end{align}
Moreover, it follows that
\begin{align}
  \text{Var}(\hat{\boldsymbol{\beta}}) & = \text{Var}(\mathbf{B}\hat{\mathbf{w}}) \\
  & = \mathbf{B} \text{Var}(\hat{\mathbf{w}}) \mathbf{B}^\top\\
  & = \mathbf{B} (-\mathbf{H}^{-1} + \boldsymbol{\Sigma}) \mathbf{B}^\top \\
  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + \mathbf{B}\boldsymbol{\Sigma}\mathbf{B}^\top \\
  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}.
\end{align}
In practice, $\text{Var}(\hat{\boldsymbol{\beta}})$ is estimated by evaluating $\boldsymbol{\Sigma}$ at $\hat{\boldsymbol{\theta}}$, the estimated covariance parameter vector.

@@

Inspecting model diagnostics is an important step of the modeling process that can yield valuable insights into model behavior and unusual observations. @montgomery2021introduction contextualize three components of unusual observations: outliers, leverage, and influence. An observation is an outlier if it has an extreme response value relative to expectation. The response GLM residuals simply compare the observation to its fitted latent mean:
\begin{align}
  \mathbf{r}_{r} = \mathbf{y} - f^{-1}(\hat{\mathbf{w}})
\end{align}
Because observations often have a unique support in a GLM (e.g., only two possible response values for binary data) and the variance of an observation generally depends on its mean, response residuals lack some utility. Deviance residuals are a function of response residuals that are appropriately scaled to behave more like response residuals in a standard linear model. Deviance residuals are given by
\begin{align}
  \mathbf{r}_{d} = sign(\mathbf{r}_{r})\sqrt{\mathbf{d}},
\end{align}
where $\mathbf{d}$ is a vector of individual deviances. The sum of the squared deviance residuals equals the sum of the elements of $\mathbf{d}$, known as the deviance of the model fit. The deviance of the model fit quantifies twice the difference in log likelihoods between the a saturated model that fits every observation perfectly (i.e., $\mathbf{y} = f^{-1}(\hat{\mathbf{w}}_i)$ for all $i$) and the fitted model [@myers2012generalized]. Deviance is often used as a fit statistic; lower values of deviance imply a better model fit (compared to the observed data). Pearson and standardized residuals are other types of GLM residuals that involve a scaling of the response residuals; the Pearson residuals scale $\mathbf{r}_{r}$ by the square root of $\mathbf{V}$, while the standardized residuals scale the deviance residuals by $\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}$, where $\mathbf{L}_{ii}$ is the $i$th diagonal element of the leverage matrix, which we discuss next.

### 2025-11-14 inst/manuscript/manuscript.tex
some updates

share spatial covariance. Following \citet{harville1977maximum}, we can

@@

\(\ell_\mathbf{w}\) around a point \(\hat{\mathbf{w}}\) yields

@@

\(\mathbf{w}\). If \(\hat{\mathbf{w}}\) is a value for which

@@

\(\exp(\ell_{\hat{\mathbf{w}}})\) yields
\begin{align}\label{eq-marginal6}
  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\hat{\mathbf{w}}), \varphi] [\hat{\mathbf{w}} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\hat{\mathbf{w}}}|^{-1/2}.

@@

Maximizing the natural logarithm of Equation\(~\)\ref{eq-marginal6}
requires a doubly iterative process over \(\boldsymbol{\theta}\) and
\(\varphi\) as well as \(\mathbf{w}\), eventually yielding the the
marginal restricted maximum likelihood estimators \(\hat{\varphi}\) and
\(\hat{\boldsymbol{\theta}}\) and their corresponding values of
\(\hat{\mathbf{w}}\). Maximizing this log likelihood requires repeatedly
evaluating \(\boldsymbol{\Sigma}^{-1}\), \(\mathbf{g}\), and
\(\mathbf{G}\); see \citet{ver2024marginal} for more details and
comparisons to other approaches as well as forms of \(\mathbf{g}\) and
\(\mathbf{G}\) for various response distributions.

@@

Though the fixed effects are integrated out of the likelihood, we can
still estimate them using generalized least squares (GLS) principles, a
common practice for linear models estimated using restricted maximum
likelihood methods. Had we observed \(\mathbf{w}\), a GLS estimator for
\(\boldsymbol{\beta}\) is given by \begin{align}\label{eq-gls1}
  \hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{w} = \mathbf{B}\mathbf{w},
\end{align} where
\(\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\).
However, we only observe \(\hat{\mathbf{w}}\), so it is reasonable to
define \(\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}\). Thus,
to derive properties of \(\hat{\boldsymbol{\beta}}\) like expectation
and variance, we must derive these properties for \(\hat{\mathbf{w}}\).
To do so, we must condition on \(\mathbf{w}\) as if it were observed and
invoke properties of the laws of total expectation and variance. Because
\(\hat{\mathbf{w}}\) was obtained by maximizing the likelihood, we may
assume that given \(\mathbf{w}\), \(\hat{\mathbf{w}}\) has mean
\(\mathbf{w}\) and variance approximately equal to \(-\mathbf{H}^{-1}\)
(the inverse Hessian). It follows that \(\text{E}(\hat{\mathbf{w}})\) is
given by \begin{align}
  \text{E}(\hat{\mathbf{w}}) = \text{E}(\text{E}(\hat{\mathbf{w}} | \mathbf{w}))
   = \text{E}(\mathbf{w}) 
   = \mathbf{X}\boldsymbol{\beta}
\end{align} and \(\text{Var}(\hat{\mathbf{w}})\) is given by
\begin{align}
  \text{Var}(\hat{\mathbf{w}}) & = \text{E}(\text{Var}(\hat{\mathbf{w}} | \mathbf{w})) + \text{Var}(\text{E}(\hat{\mathbf{w}} | \mathbf{w})) \\
  & = \text{E}(-\mathbf{H}^{-1}) + \text{Var}(\mathbf{w})\\
  & = -\mathbf{H}^{-1} + \boldsymbol{\Sigma}
\end{align} Putting this all together, it follows that
\(\hat{\boldsymbol{\beta}}\) is unbiased for \(\boldsymbol{\beta}\):
\begin{align}
  \text{E}(\hat{\boldsymbol{\beta}})  = \text{E}(\mathbf{B}\hat{\mathbf{w}}) 
  = \mathbf{B}\text{E}(\hat{\mathbf{w}}) =  (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})\boldsymbol{\beta} = \boldsymbol{\beta}.
\end{align} Moreover, it follows that \begin{align}
  \text{Var}(\hat{\boldsymbol{\beta}}) & = \text{Var}(\mathbf{B}\hat{\mathbf{w}}) \\
  & = \mathbf{B} \text{Var}(\hat{\mathbf{w}}) \mathbf{B}^\top\\
  & = \mathbf{B} (-\mathbf{H}^{-1} + \boldsymbol{\Sigma}) \mathbf{B}^\top \\
  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + \mathbf{B}\boldsymbol{\Sigma}\mathbf{B}^\top \\
  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}.
\end{align} In practice, \(\text{Var}(\hat{\boldsymbol{\beta}})\) is
estimated by evaluating \(\boldsymbol{\Sigma}\) at
\(\hat{\boldsymbol{\theta}}\), the estimated covariance parameter
vector.

@@

Inspecting model diagnostics is an important step of the modeling
process that can yield valuable insights into model behavior and unusual
observations. \citet{montgomery2021introduction} contextualize three
components of unusual observations: outliers, leverage, and influence.
An observation is an outlier if it has an extreme response value
relative to expectation. The response GLM residuals simply compare the
observation to its fitted latent mean: \begin{align}
  \mathbf{r}_{r} = \mathbf{y} - f^{-1}(\hat{\mathbf{w}})
\end{align} Because observations often have a unique support in a GLM
(e.g., only two possible response values for binary data) and the
variance of an observation generally depends on its mean, response
residuals lack some utility. Deviance residuals are a function of
response residuals that are appropriately scaled to behave more like
response residuals in a standard linear model. Deviance residuals are
given by \begin{align}
  \mathbf{r}_{d} = sign(\mathbf{r}_{r})\sqrt{\mathbf{d}},
\end{align} where \(\mathbf{d}\) is a vector of individual deviances.
The sum of the squared deviance residuals equals the sum of
\(\mathbf{d}\). The sum of \(\mathbf{d}\) is the deviance of the model
fit, which quantifies twice the difference in log likelihoods between
the a saturated model that fits every observation perfectly (i.e.,
\(\mathbf{y} = f^{-1}(\hat{\mathbf{w}}_i)\) for all \(i\)) and the
fitted model \citep{myers2012generalized}. Deviance is often used as a
fit statistic; lower values of deviance imply a better model fit
(compared to the observed data). Pearson and standardized residuals are
other types of GLM residuals that involve a scaling of the response
residuals; the Pearson residuals scale \(\mathbf{r}_{r}\) by the square
root of \(\mathbf{V}\), while the standardized residuals scale the
deviance residuals by \(\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}\), where
\(\mathbf{L}_{ii}\) is the \(i\)th diagonal element of the leverage
matrix, which we discuss next.

@@

We would like to the associate editor, anonymous reviewers, and
editorial staff for significant support and feedback that greatly
improved the manuscript.

>>>>

share spatial covariance. Following \citet{harville1974bayesian}, we can

@@

\(\ell_\mathbf{w}\) around a point \(\mathbf{w}^{*}\) yields

@@

\(\mathbf{w}\). If \(\mathbf{w}^{*}\) is a value for which

@@

\(\exp(\ell_{\mathbf{w}^{*}})\) yields \begin{align}\label{eq-marginal6}
  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\mathbf{w}^{*}), \varphi] [\mathbf{w}^{*} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\mathbf{w}^{*}}|^{-1/2}.

@@

Maximizing the natural logarithm of Equation\(~\)\ref{eq-marginal6}
requires a doubly iterative process over 1) \(\boldsymbol{\theta}\) and
\(\varphi\) and 2) the latent \(\mathbf{w}\) (to find each set of
\(\mathbf{w}^{*}\)). This doubly iterative process eventually yields the
marginal restricted maximum likelihood estimators \(\hat{\varphi}\) and
\(\hat{\boldsymbol{\theta}}\) and their respective values of
\(\mathbf{w}^{*}\), which we call \(\hat{\mathbf{w}}\). Maximizing the
marginal restricted likelihood requires repeatedly evaluating
\(\boldsymbol{\Sigma}^{-1}\), \(\mathbf{g}\), and \(\mathbf{G}\); see
\citet{ver2024marginal} for more details and comparisons to other
approaches as well as forms of \(\mathbf{g}\) and \(\mathbf{G}\) for
various response distributions.

@@

Though the fixed effects are integrated out of the likelihood, we can
still estimate them using generalized least squares (GLS) principles, a
common practice for linear models estimated using restricted maximum
likelihood methods. Had we observed \(\mathbf{w}\), a GLS estimator for
\(\boldsymbol{\beta}\) is given by \begin{align}\label{eq-gls1}
  \hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{w} = \mathbf{B}\mathbf{w},
\end{align} where
\(\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\).
However, we do not observe \(\mathbf{w}\) and instead estimate
\(\mathbf{w}\) via \(\hat{\mathbf{w}}\). Thus, it is reasonable to
define \(\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}\). To
derive properties of \(\hat{\boldsymbol{\beta}}\) like expectation and
variance, we must derive these properties for \(\hat{\mathbf{w}}\). To
do so, we must condition on \(\mathbf{w}\) as if it were observed and
invoke properties of the laws of total expectation and variance. Because
\(\hat{\mathbf{w}}\) was obtained by maximizing the likelihood, we may
assume that given \(\mathbf{w}\), \(\hat{\mathbf{w}}\) has mean
\(\mathbf{w}\) and variance approximately equal to \(-\mathbf{H}^{-1}\)
(the inverse Hessian). It follows that \(\text{E}(\hat{\mathbf{w}})\) is
given by \begin{align}
  \text{E}(\hat{\mathbf{w}}) = \text{E}(\text{E}(\hat{\mathbf{w}} | \mathbf{w}))
   = \text{E}(\mathbf{w}) 
   = \mathbf{X}\boldsymbol{\beta}
\end{align} and \(\text{Var}(\hat{\mathbf{w}})\) is given by
\begin{align}
  \text{Var}(\hat{\mathbf{w}}) & = \text{E}(\text{Var}(\hat{\mathbf{w}} | \mathbf{w})) + \text{Var}(\text{E}(\hat{\mathbf{w}} | \mathbf{w})) \\
  & = \text{E}(-\mathbf{H}^{-1}) + \text{Var}(\mathbf{w})\\
  & = -\mathbf{H}^{-1} + \boldsymbol{\Sigma}
\end{align} Putting this all together, it follows that
\(\hat{\boldsymbol{\beta}}\) is unbiased for \(\boldsymbol{\beta}\):
\begin{align}
  \text{E}(\hat{\boldsymbol{\beta}})  = \text{E}(\mathbf{B}\hat{\mathbf{w}}) 
  = \mathbf{B}\text{E}(\hat{\mathbf{w}}) =  (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})\boldsymbol{\beta} = \boldsymbol{\beta}.
\end{align} Moreover, it follows that \begin{align}
  \text{Var}(\hat{\boldsymbol{\beta}}) & = \text{Var}(\mathbf{B}\hat{\mathbf{w}}) \\
  & = \mathbf{B} \text{Var}(\hat{\mathbf{w}}) \mathbf{B}^\top\\
  & = \mathbf{B} (-\mathbf{H}^{-1} + \boldsymbol{\Sigma}) \mathbf{B}^\top \\
  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + \mathbf{B}\boldsymbol{\Sigma}\mathbf{B}^\top \\
  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}.
\end{align} In practice, \(\text{Var}(\hat{\boldsymbol{\beta}})\) is
estimated by evaluating \(\boldsymbol{\Sigma}\) at
\(\hat{\boldsymbol{\theta}}\), the estimated covariance parameter
vector.

@@

Inspecting model diagnostics is an important step of the modeling
process that can yield valuable insights into model behavior and unusual
observations. \citet{montgomery2021introduction} contextualize three
components of unusual observations: outliers, leverage, and influence.
An observation is an outlier if it has an extreme response value
relative to expectation. The response GLM residuals simply compare the
observation to its fitted latent mean: \begin{align}
  \mathbf{r}_{r} = \mathbf{y} - f^{-1}(\hat{\mathbf{w}})
\end{align} Because observations often have a unique support in a GLM
(e.g., only two possible response values for binary data) and the
variance of an observation generally depends on its mean, response
residuals lack some utility. Deviance residuals are a function of
response residuals that are appropriately scaled to behave more like
response residuals in a standard linear model. Deviance residuals are
given by \begin{align}
  \mathbf{r}_{d} = sign(\mathbf{r}_{r})\sqrt{\mathbf{d}},
\end{align} where \(\mathbf{d}\) is a vector of individual deviances.
The sum of the squared deviance residuals equals the sum of the elements
of \(\mathbf{d}\), known as the deviance of the model fit. The deviance
of the model fit quantifies twice the difference in log likelihoods
between the a saturated model that fits every observation perfectly
(i.e., \(\mathbf{y} = f^{-1}(\hat{\mathbf{w}}_i)\) for all \(i\)) and
the fitted model \citep{myers2012generalized}. Deviance is often used as
a fit statistic; lower values of deviance imply a better model fit
(compared to the observed data). Pearson and standardized residuals are
other types of GLM residuals that involve a scaling of the response
residuals; the Pearson residuals scale \(\mathbf{r}_{r}\) by the square
root of \(\mathbf{V}\), while the standardized residuals scale the
deviance residuals by \(\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}\), where
\(\mathbf{L}_{ii}\) is the \(i\)th diagonal element of the leverage
matrix, which we discuss next.

@@

We would like to genuinely thank the associate editor, anonymous
reviewers, and editorial staff for significant support and feedback that
greatly improved the manuscript.
